---
title: 'malbacR: A Vignette Pertaining to Pre-Filtered Data'
author: "Damon Leach, Kelly Stratton, Lisa Bramer"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    rmarkdown::html_vignette:
    fig_caption: yes
    self_contained: yes
    toc: true
bibliography: analysis_ref.bib
vignette: >
  %\VignetteIndexEntry{malbacR-condensed-filtered} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

The R package `malbacR` is a new package that deals with batch correction methods of small molecule omics data. It works in conjunction with `pmartR`, an R package designed for preprocessing, filtering, and analyzing multi-omics data [@pmartr; @pmartr2].

This outline demonstrates the functionality and use of the `malbacR` package. Within this package, there are four data sets: `pmart_amide`, `pmart_amideFilt`,`pmart_mix`, and `pmart_mixFilt`. The first two objects contain data originally found in the package `WaveICA2.0` [@waveica_package]. The second two objects contain data originally found in the package `crmn` [@crmn_package]. The data sets with the suffix "Filt" are filtered versions of the original data sets. That is, unlike their non-filtered counterparts, these data objects require no additional data manipulation for batch correction to successfully run. For this example, we work with the filtered data sets to demonstrate how the `malbacR` functions work.

Within `malbacR` there are 12 batch correction methods:

1. Range Scaling [@scaling_paper]
2. Power Scaling [@scaling_paper]
3. Pareto Scaling [@scaling_paper]
4. ComBat [@combat_paper]
5. EigenMS [@eigenMS_paper]
6. NOMIS [@nomis_paper]
7. RUV-random [@ruvrandom_paper]
8. QC-RLSC [@qcrlsc_paper]
9. WaveICA2.0 [@waveica_paper]
10. TIGER [@tiger_paper]
11. SERRF [@serrf_paper]
12. QC-RFSC [@qcrfsc_paper]

```{r, echo = FALSE}
knitr::opts_chunk$set(message = FALSE)
```

```{r}
library(malbacR)
library(pmartR)
library(ggplot2)
```

## Data Set 1: Amide Data

### Load in the Data

A `pmartR` friendly version of the "Amide" data that has already undergone filtering and log2 transformations is already implemented within `malbacR`. Therefore, we simply load in the data.

```{r}
data(pmart_amideFilt)
```

### Run Batch Correction Methods

The "Amide" data set contains information regarding quality control samples. Therefore the following methods can be used:

- Range Scaling
- Power Scaling
- Pareto Scaling
- TIGER
- EigenMS
- WaveICA2.0
- SERRF
- ComBat
- QCRLSC
- QCRFSC

It is important to note that the data `pmart_amideFilt` has been filtered, transformed, and imputed such that the data runs with all possible methods, but it may be conservative for a given method. For example, some methods do not require imputation to run, but because some methods require imputation (like SERRF and WaveICA2.0) [@serrf_paper;@waveica_paper], the data has been imputed.

The data is currently on a log2 scale, therefore, the data must be converted to abundance values for the following methods: SERRF, TIGER, QC-RFSC, and WaveICA.

```{r, results = 'hide'}
# SCALING METHODS
# range scaling
amide_range <- bc_range(omicsData = pmart_amideFilt)

# power scaling
amide_power <- bc_power(omicsData = pmart_amideFilt)

# pareto scaling
amide_pareto <- bc_pareto(omicsData = pmart_amideFilt)

# QUALITY CONTROL METHODS
# TIGER
pmart_amideFilt_abundance <- edata_transform(omicsData = pmart_amideFilt, data_scale = "abundance")
amide_tiger_abundance <- bc_tiger(omicsData = pmart_amideFilt_abundance,sampletype_cname = "group",
                        test_val = "QC",injection_cname = "Injection_order",group_cname = "group")
amide_tiger <- edata_transform(omicsData = amide_tiger_abundance, data_scale = "log2")

# QC-RLSC
amide_qcrlsc <- bc_qcrlsc(omicsData = pmart_amideFilt,block_cname = "batch",
                          qc_cname = "group", qc_val = "QC", order_cname = "Injection_order",
                          missing_thresh = 0.5, rsd_thresh = 0.3, backtransform  = FALSE,keep_qc = FALSE)

# SERRF
amide_serrf_abundance <- bc_serrf(omicsData = pmart_amideFilt_abundance,sampletype_cname = "group",test_val = "QC",group_cname = "group")
amide_serrf <- edata_transform(omicsData = amide_serrf_abundance, data_scale = "log2")

# OTHER METHODS
# ComBat
amide_combat <- bc_combat(omicsData = pmart_amideFilt, use_groups = FALSE)

# EigenMS
amide_eigen <- bc_eigenMS(omicsData = pmart_amideFilt)

# WaveICA2.0
amide_wave_abundance <- bc_waveica(omicsData = pmart_amideFilt_abundance, injection_cname = "Injection_order",
                         version = "WaveICA2.0", alpha = 0, cutoff_injection = 0.1, K = 10,
                         negative_to_na = TRUE)
amide_wave <- edata_transform(omicsData = amide_wave_abundance, data_scale = "log2")

# QC-RFSC
amide_qcrfsc_abundance <- bc_qcrfsc(omicsData = pmart_amideFilt_abundance,qc_cname = "group",qc_val = "QC",
                          order_cname = "Injection_order",ntree = 500, keep_qc = FALSE,group_cname = "group")
amide_qcrfsc <- edata_transform(amide_qcrfsc_abundance, data_scale = "log2")
```

### Data Visualization

After obtaining all the different batch corrected data sets, we can plot the PPCA to see if they are successfully returning batch corrected data. We set the `group_designation` to use our batch information as the `main_effects` so as to color the data by batch. All of this code is run using functions from `pmartR` demonstrating the utility between the two packages.

```{r, out.width = "33%"}

p1 <- plot(dim_reduction(omicsData = pmart_amideFilt),omicsData = pmart_amide,color_by = "batch") + labs(title = "Amide: Unadjusted")
p2 <- plot(dim_reduction(omicsData = amide_range),omicsData = amide_range,color_by = "batch") + labs(title = "Amide: Range")
p3 <- plot(dim_reduction(omicsData = amide_power),omicsData = amide_power,color_by = "batch") + labs(title = "Amide: Power")
p4 <- plot(dim_reduction(omicsData = amide_pareto),omicsData = amide_pareto,color_by = "batch") + labs(title = "Amide: Pareto")
p5 <- plot(dim_reduction(omicsData = amide_tiger),omicsData = amide_tiger,color_by = "batch") + labs(title = "Amide: TIGER")
p6 <- plot(dim_reduction(omicsData = amide_qcrlsc),omicsData = amide_qcrlsc,color_by = "batch") + labs(title = "Amide: QC-RLSC")
p7 <- plot(dim_reduction(omicsData = amide_combat),omicsData = amide_combat,color_by = "batch") + labs(title = "Amide: ComBat")
p8 <- plot(dim_reduction(omicsData = amide_wave),omicsData = amide_wave,color_by = "batch") + labs(title = "Amide: WaveICA2.0")
p9 <- plot(dim_reduction(omicsData = amide_serrf),omicsData = amide_serrf,color_by = "batch") + labs(title = "Amide: SERRF")
p10 <- plot(dim_reduction(omicsData = amide_qcrfsc),omicsData = amide_qcrfsc,color_by = "batch") + labs(title = "Amide: QC-RFSC")
p11 <- plot(dim_reduction(omicsData = amide_eigen),omicsData = amide_eigen,color_by = "batch") + labs(title = "Amide: EigenMS")

p1;p2;p3;p4;p5;p6;p7;p8;p9;p10;p11
```

## Data Set 2: Mix Data

### Load in the Data

The `malbacR` package includes another `pmartR` friendly data set for the "mix" data which has already undergone filtering and log2 transformations is already implemented within `malbacR`. Therefore, we simply load in the data.

```{r}
data(pmart_mixFilt)
```

### Run Batch Correction Methods

The "mix" data set contains information regarding negative controls/internal standards. Therefore the following methods can be used:

- Range Scaling
- Power Scaling
- Pareto Scaling
- RUV-random
- NOMIS
- ComBat

As there is no information regarding QC samples or injection order, the other batch correction methods in the package or not able to be used with this data set.

As with the `pmart_amideFilt` data set, is important to note that the data in `pmart_mixFilt` has been filtered and transformed such that the data runs with all possible methods, but it may be conservative for a given method. As there were no missing data with the "mix" data set, no imputation was computed.

The data is currently on a log2 scale, therefore, the data must be converted to abundance values for the following methods: NOMIS and WaveICA.

```{r, warning = FALSE}
# SCALING METHODS
# range scaling
mix_range <- bc_range(omicsData = pmart_mixFilt)

# pareto scaling
mix_pareto <- bc_pareto(omicsData = pmart_mixFilt)

# power scaling
mix_power <- bc_power(omicsData = pmart_mixFilt)

# INTERNAL STANDARDS/NEGATIVE CONTROLS
# RUV-random
mix_ruv <- bc_ruvRandom(omicsData = pmart_mixFilt, nc_cname = "tag",nc_val = "IS", k = 3)

# NOMIS
pmart_mixFilt_abundance <- edata_transform(omicsData = pmart_mixFilt, data_scale = "abundance")
mix_nomis_abundance <- bc_nomis(omicsData = pmart_mixFilt_abundance, is_cname = "tag", is_val = "IS")
mix_nomis <- edata_transform(omicsData = mix_nomis_abundance, data_scale = "log2")

# OTHER METHODS
mix_combat <- bc_combat(omicsData = pmart_mixFilt)

mix_waveica_abundance <- bc_waveica(omicsData = pmart_mixFilt_abundance,batch_cname = "BatchNum",
                                    version = "WaveICA",alpha = 0, K = 20, cutoff_batch = 0.05, 
                                    cutoff_group = 0.05,negative_to_na = TRUE)
mix_waveica <- edata_transform(omicsData = mix_waveica_abundance, data_scale = "log2")
```

### Mix: Data Visualization

Similar to the previous data set, we can compare the PCA plots between the unadjusted and adjusted data sets.

```{r,out.width = "33%"}
p1 <- plot(dim_reduction(omicsData = pmart_mixFilt),omicsData = pmart_mixFilt,color_by = "BatchNum") + labs(title = "Mix: Unadjusted")
p2 <- plot(dim_reduction(omicsData = mix_ruv),omicsData = mix_ruv,color_by = "BatchNum") + labs(title = "Mix: ruv-Random")
p3 <- plot(dim_reduction(omicsData = mix_nomis),omicsData = mix_nomis,color_by = "BatchNum") + labs(title = "Mix: NOMIS")
p4 <- plot(dim_reduction(omicsData = mix_combat),omicsData = mix_combat,color_by = "BatchNum") + labs(title = "Mix: ComBat")
p5 <- plot(dim_reduction(omicsData = mix_range),omicsData = mix_range,color_by = "BatchNum") + labs(title = "Mix: Range")
p6 <- plot(dim_reduction(omicsData = mix_power),omicsData = mix_power,color_by = "BatchNum") + labs(title = "Mix: Power")
p7 <- plot(dim_reduction(omicsData = mix_pareto),omicsData = mix_pareto,color_by = "BatchNum") + labs(title = "Mix: Pareto")
p8 <- plot(dim_reduction(omicsData = mix_waveica),omicsData = mix_waveica,color_by = "BatchNum") + labs(title = "Mix: WaveICA")

p1;p2;p3;p4;p5;p6;p7;p8
```

## References
